---
title: "Past Data Visualizations"
description: |
  Some past visualizations I made in Math 230.
author:
  - name: Mikey Myro
date: "May 4,  2026"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(ggplot2)
library(aRtsy)
library(palmerpenguins)
library(gapminder)
```

# Past Course Visualizations

## Night Motion

```{r}
set.seed(15)
canvas_strokes(colors = c("black", "grey40", "grey70", "white"))
```

## Sky Flow

```{r}
set.seed(150)
canvas_ribbons(colors = colorPalette("sky"))
```

## Bucknell Spirit

```{r}
set.seed(5)
canvas_tiles(colors = c("#FF6600", "#0033A0", "#FFFFFF", "#CCCCCC"))
```

## Flipper Length vs Bill Length

```{r}
penguins <- na.omit(palmerpenguins::penguins)

ggplot(penguins, aes(x = bill_length_mm, y = flipper_length_mm)) +
  geom_point(aes(color = island, shape = species)) +
  geom_smooth(method = "loess") +
  labs(
    title = "Flipper Length vs Bill Length",
    x = "Bill Length (mm)",
    y = "Flipper Length (mm)",
    color = "Island",
    shape = "Species"
  ) +
  theme_minimal()
```

**Data/Variables:** Uses the Palmer Penguins dataset with bill length, flipper length, species, and island.

**What it shows:** As bill length increases, flipper length also tends to increase, with visible differences across species.

## GDP vs Life Expectancy

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.4) +
  geom_density_2d() +
  scale_x_log10() +
  labs(
    title = "GDP vs Life Expectancy",
    x = "GDP per Capita",
    y = "Life Expectancy"
  ) +
  theme_minimal()
```

**Data/Variables:** Uses the Gapminder dataset with GDP per capita and life expectancy.

**What it shows:** Countries with lower GDP tend to have lower life expectancy, and life expectancy increases and levels off as GDP rises.

## Confidence Interval Dashboard


```{r}
set.seed(1)

x <- 1:50
y <- 2*x + rnorm(50, 0, 10)

df <- data.frame(x, y)

model <- lm(y ~ x, data = df)

ci95 <- as.data.frame(predict(model, interval = "confidence", level = 0.95))

plot_df <- data.frame(
  x = df$x,
  y = df$y,
  fit = ci95$fit,
  lwr = ci95$lwr,
  upr = ci95$upr
)

ggplot(plot_df, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = fit), color = "black") +
  geom_line(aes(y = lwr), linetype = "dashed", color = "red") +
  geom_line(aes(y = upr), linetype = "dashed", color = "red") +
  labs(
    title = "Linear Fit with 95% Confidence Interval",
    x = "X",
    y = "Y"
  ) +
  theme_minimal()
```

**Data/Variables:** Simulated data used to demonstrate confidence intervals.

**What it shows:** Higher confidence levels produce wider intervals, while larger sample sizes make intervals narrower.