---
title: "Run Fastballs"
description: |
  How extreme horizontal movement changes the shape and effectiveness of a four-seam fastball.
author:
  - name: Mikey Myro
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    css: fastball_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)

source("fastball_setup.R")
```

<style>

.analysis-box {
  background: #f7f7f7;
  border-left: 5px solid #009E73;
  border-radius: 8px;
  padding: 20px 22px;
  margin: 22px 0 30px 0;
  font-size: 15px;
  line-height: 1.7;
}

.player-grid {
  display: grid;
  grid-template-columns: repeat(2, 1fr);
  gap: 16px;
  margin: 22px 0 32px 0;
}

.player-card {
  border: 1px solid #e3e3e3;
  border-radius: 10px;
  padding: 18px 20px;
  background: #fafafa;
}

.player-name {
  font-size: 19px;
  font-weight: 700;
  margin-bottom: 6px;
}

.player-label {
  font-size: 12px;
  font-weight: 700;
  color: #009E73;
  text-transform: uppercase;
  letter-spacing: 0.5px;
  margin-bottom: 8px;
}

.player-text {
  font-size: 14px;
  line-height: 1.6;
  color: #444444;
}

.comparison-intro {
  font-size: 15px;
  color: #555555;
  line-height: 1.6;
  margin: 15px 0 20px 0;
}

.disclaimer-box {
  background: #f7f7f7;
  border: 1px solid #dedede;
  border-top: 5px solid #009E73;
  border-radius: 10px;
  padding: 22px 24px;
  margin: 24px 0 32px 0;
  line-height: 1.7;
}

@media (max-width: 750px) {
  .player-grid {
    grid-template-columns: 1fr;
  }
}

</style>

<div class="profile-hero run-hero">

<div class="hero-title">Run</div>

<div class="hero-subtitle">

Four-seam fastballs with more horizontal arm-side movement than comparable pitches. These fastballs create a very different shape from the traditional high-ride four-seamer.

</div>

</div>

# Run

The pitch-classification disclaimer matters more on this page than anywhere else.

These are <strong>running four-seam fastballs, not sinkers</strong>.

A lot of the most extreme arm-side fastballs in baseball are classified by Baseball Savant as sinkers, so they never enter this group. This is not a full evaluation of whether run itself is good or bad.

It is specifically a look at what happens when pitches that are still classified as four-seam fastballs have much more horizontal movement than comparable four-seamers.

<div class="disclaimer-box">

<strong>Pitch Classification Matters</strong>

<br><br>

Many pitchers who intentionally chase extreme arm-side movement throw pitches that are classified as sinkers rather than four-seam fastballs.

That means this group should not be viewed as the best running fastballs in baseball. It represents the four-seamers that sit on the extreme running side of the four-seam movement spectrum.

</div>

<div class="info-grid">

<div class="info-card">
<div class="info-title">Ranking</div>
<div class="info-value">HB vs Comparable</div>
<div class="info-description">

The 15 four-seamers with the most horizontal arm-side movement relative to comparable pitches.

</div>
</div>

<div class="info-card">
<div class="info-title">Physical Trait</div>
<div class="info-value">Arm-Side Movement</div>
<div class="info-description">

The farther the pitch moves horizontally toward the arm side relative to similar fastballs, the more run it has.

</div>
</div>

<div class="info-card">
<div class="info-title">Classification</div>
<div class="info-value">Four-Seam Only</div>
<div class="info-description">

Extreme running fastballs are often classified as sinkers, so this group only captures one part of the overall run spectrum.

</div>
</div>

</div>

## Most Horizontal Run

<div class="section-intro">

These were the 15 four-seam fastballs with the most horizontal movement relative to comparable pitches.

</div>

```{r fig.width=9, fig.height=6}
run_page$leaderboard
```

# Movement Profiles

Horizontal run can be influenced by several parts of a pitcher’s delivery and pitch profile, especially spin direction and release characteristics.

Arm slot is one of the easiest examples to see. A lower arm slot often comes with a more horizontal spin direction, which can contribute to additional arm-side movement.

It is not as simple as saying lower slot automatically equals more run, but the relationship between arm slot, spin axis, and horizontal movement is important.

Jake Diekman is a good example. His extremely low arm slot gives hitters a completely different look, and his four-seam fastball produced far more horizontal movement than comparable pitches.

The vertical component matters too.

Lower IVB combined with a lot of arm-side movement begins to look more like the shape normally associated with a sinker. A four-seam sitting around the mid-teens in IVB while also producing significant run can sometimes fall into what pitchers and coaches describe as a flatter fastball profile rather than a true riding fastball.

That is part of what makes this group interesting. A lot of these pitches sit near the boundary between a traditional four-seam shape and pitches that could easily be classified differently if the movement became slightly more extreme.

## Actual Movement

<div class="section-intro">

The raw movement chart shows how much variation still exists inside a group defined by run. Some pitches combine arm-side movement with lower IVB, while others still maintain more vertical ride.

</div>

```{r fig.width=9, fig.height=7}
run_page$movement
```

## Movement Relative to Comparable Fastballs

<div class="section-intro">

Moving farther to the right represents more horizontal run relative to comparable four-seamers.

The vertical axis helps show whether that run is paired with ride, average IVB, or a flatter and more sinker-like vertical shape.

</div>

```{r fig.width=9, fig.height=7}
run_page$profile_space
```

<div class="analysis-box">

<strong>Run by itself does not completely describe the pitch.</strong>

A high-run fastball with strong IVB can still look very different from a high-run fastball with lower IVB. The vertical component changes whether the pitch behaves more like a riding four-seamer, a flatter fastball, or begins approaching the movement profile normally associated with a sinker.

That is why the full movement profile matters more than simply saying a pitch has a lot of horizontal break.

</div>

<div class="player-grid">

<div class="player-card">

<div class="player-label">Extreme Run</div>

<div class="player-name">Jake Diekman</div>

<div class="player-text">

Diekman is one of the clearest examples of how delivery characteristics can contribute to an unusual fastball shape.

His low arm slot creates a very different release look, and his four-seam produced substantially more horizontal movement than comparable pitches.

</div>

</div>

<div class="player-card">

<div class="player-label">Run and Vertical Shape</div>

<div class="player-name">The Sinker Boundary</div>

<div class="player-text">

As horizontal run increases and IVB decreases, the pitch begins to approach the movement profile normally associated with a sinker.

That makes the Run group especially sensitive to how Baseball Savant classifies each pitch.

</div>

</div>

</div>

# Run and Expected Results

<div class="section-intro">

Comparing horizontal movement relative to comparable fastballs with expected wOBA shows whether additional run consistently translated into better results.

</div>

```{r fig.width=9, fig.height=7}
run_page$relationship
```

# Effectiveness

The results are not especially strong for the high-run four-seam group.

Only five of the 15 fastballs finished with a positive Run Value per 100, and Richard Rodríguez led the group at only around 1.0 RV/100.

Compared with the other fastball shapes, this was not the most effective group.

That does not mean arm-side run itself is bad. Many pitchers who intentionally chase large amounts of arm-side movement are throwing sinkers, and those pitches are completely absent from this sample.

What it does suggest is that extreme run on a pitch that was still being thrown and classified as a four-seamer was not especially effective within this 2021 group.

## Expected wOBA

Lower xwOBA shows which high-run fastballs were best at limiting expected offensive production.

```{r fig.width=9, fig.height=6}
run_page$xwoba
```

## Whiff Rate

Whiff rate shows whether the additional horizontal movement actually translated into more swings and misses.

```{r fig.width=9, fig.height=6}
run_page$whiff
```

## Run Value per 100

This is where the weakness of the group becomes especially clear. Only a small number of these high-run four-seamers produced positive value on a per-pitch basis.

```{r fig.width=9, fig.height=6}
run_page$run_value
```

## Hard-Hit Rate

Hard-Hit Rate provides another way to see whether the movement helped prevent hitters from squaring the pitch up when they did make contact.

```{r fig.width=9, fig.height=6}
run_page$hard_hit
```

# Individual Results

<div class="player-grid">

<div class="player-card">

<div class="player-label">Lowest Results in the Group</div>

<div class="player-name">Jesús Luzardo</div>

<div class="player-text">

Luzardo had one of the worst four-seam fastballs in the group that season.

He has developed into a much better pitcher since then, but 2021 was a terrible season for him and he was still far from the pitcher he would eventually become.

His appearance near the bottom of this group is a good reminder that a pitcher should be evaluated based on the pitch and season in front of you rather than his reputation or what he becomes later.

</div>

</div>

<div class="player-card">

<div class="player-label">Established Pitcher, Down Season</div>

<div class="player-name">Aaron Nola</div>

<div class="player-text">

Nola has been one of the better starting pitchers of the last decade, but 2021 was a poor season by his standards.

His four-seam was slightly positive by RV/100, but it was nowhere near a dominant pitch.

A recognizable name does not automatically mean every pitch in every season was effective.

</div>

</div>

</div>

<div class="analysis-box">

Overall, high horizontal run was the least convincing four-seam shape in this data.

The group did not consistently miss bats at an elite level, did not dominate expected results, and produced relatively weak Run Value per 100 numbers.

I would not use that as evidence that running fastballs are ineffective.

If anything, it may suggest that when a pitcher is trying to maximize arm-side movement, the better version of that pitch often becomes a <strong>sinker rather than a four-seamer</strong>.

The four-seam classification itself is an important part of the result.

</div>

# Run Group vs Qualified Average

<div class="comparison-intro">

The group average makes it easier to compare the 15 highest-run four-seam fastballs with the qualified four-seam population as a whole.

</div>

```{r}
make_effectiveness_comparison(
  run_fastballs,
  "Top 15 Run Fastballs"
)
```

# Complete Profiles

<div class="takeaway-box">

<div class="takeaway-title">Run in Context</div>

<div class="takeaway-text">

Horizontal movement can create a unique fastball shape, but the Run group shows that unusual movement alone does not guarantee effectiveness.

The most important context is that these are still four-seam fastballs. Many of the pitches that maximize arm-side movement are classified as sinkers and therefore fall outside this group entirely.

</div>

</div>

```{r}
make_stats_table(
  run_fastballs,
  "Top 15 Four-Seam Fastballs by Horizontal Break vs Comparable"
)
```